---
title: "Biostat 203B Homework 4"
subtitle: Due Mar 18 @ 11:59PM
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

Display machine information:
```{r}
sessionInfo()
```
Load database libraries and the tidyverse frontend:
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(miceRanger))
```

## Q1. Missing data

Through the Shiny app developed in HW3, we observe abundant missing values in the MIMIC-IV ICU cohort we created. In this question, we use multiple imputation to obtain a data set without missing values.

0. Read following tutorials on the R package miceRanger for imputation: <https://github.com/farrellday/miceRanger>, <https://cran.r-project.org/web/packages/miceRanger/vignettes/miceAlgorithm.html>.

    A more thorough book treatment of the practical imputation strategies is the book [*_Flexible Imputation of Missing Data_*](https://stefvanbuuren.name/fimd/) by Stef van Buuren. 

1. Explain the jargon MCAR, MAR, and MNAR.
## Solution

- MCAR stands for "missing completely at rondom." This means that the cause of missing data is unrelated to the data itself, so there are no systematic differences between observations with and without missing data. Therefore, no bias is introduced, but the statistical power will be lower.

- MAR stands for "missing at random." This means that missing data is systematically different from the observed data, but this difference can be explained by other fully observed variables.

- MNAR stands for "missing not at random." This means that missing data is systematically related to unobserved data that are not measured by the researcher.

2. Explain in a couple of sentences how the Multiple Imputation by Chained Equations (MICE) work.

- MICE is a multiple imputation method that replaces missing data under assumptions about the missingness mechanism (MCAR, MAR, or MNAR). MICE fits a separate regression model for each variable with missing data based on the other variables to predict missing values. In order to fit the regression model, temporary placeholders are inputted for all variables with missing data. When MICE reaches a variable, all of its placeholders are set back to missing when the regression is conducted to generate predicted values. This process is repeated for all such variables a specified number of times (i.e., number of cycles).

3. Perform a data quality check of the ICU stays data. Discard variables with substantial missingness, say >5000 `NA`s. Replace apparent data entry errors by `NA`s.

## Solution

```{r}
# Load the Data
icu_cohort <- readRDS("icu_cohort.rds")
# Check for substantial missingness
summary(icu_cohort)
# Discard deathtime, edregtime, edouttime, and dod (>5000 NAs)
icu_cohort <- select(icu_cohort, -c(deathtime, edregtime, edouttime, dod))
# Remove outliers using 1.5 IQR rule
summary(icu_cohort)
for (i in 23:38) {
  col_vec <- icu_cohort[[i]]
  iqr <- IQR(col_vec, na.rm = TRUE)
  q1 <- quantile(col_vec, 0.25, na.rm = TRUE)
  q3 <- quantile(col_vec, 0.75, na.rm = TRUE)
  col_vec[col_vec > q3 + 1.5 * iqr] <- NA
  col_vec[col_vec < q1 - 1.5 * iqr] <- NA
  icu_cohort[[i]] <- col_vec
}
# Remove variables not used in Question 2
icu_cohort_final <- select(icu_cohort, -c(1:9, 12:18, 20:22))
icu_cohort_final
```

4. Impute missing values by `miceRanger` (request $m=3$ data sets). This step is computational intensive. Make sure to save the imputation results as a file. Hint: Setting `max.depth=10` in the `miceRanger` function may cut some computing time.

## Solution
```{r}
mice <- miceRanger(
  data = icu_cohort_final,
  m = 3,
  max.depth = 10,
  returnModels = FALSE,
  verbose = FALSE
)
saveRDS(mice, "./mice.rds")
```

5. Make imputation diagnostic plots and explain what they mean.

6. Choose one of the imputed data sets to be used in Q2. This is **not** a good idea to use just one imputed data set or to average multiple imputed data sets. Explain in a couple of sentences what the correct Multiple Imputation strategy is.

## Q2. Predicting 30-day mortality

Develop at least two analytic approaches for predicting the 30-day mortality of patients admitted to ICU using demographic information (gender, age, marital status, ethnicity), first lab measurements during ICU stay, and first vital measurements during ICU stay. For example, you can use (1) logistic regression (`glm()` function in base R or keras), (2) logistic regression with lasso penalty (glmnet or keras package), (3) random forest (randomForest package), or (4) neural network (keras package).

1. Partition data into 80% training set and 20% test set. Stratify partitioning according the 30-day mortality status.

2. Train the models using the training set.

3. Compare model prediction performance on the test set.

