---
title: "Biostat 203B Homework 1"
author: Jonathan King
subtitle: Due Jan 21 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=T}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email. You`ll get GitHub Pro account for free (unlimited public and private repositories).

2. Create a **private** repository `biostat-203b-2022-winter` and add `Hua-Zhou` and `maschepps` as your collaborators with write permission.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `main` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `main` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and extra data sets to reproduce results) in `main` branch.

4. After each homework due date, teaching assistant and instructor will check out your main branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

## Q2. Data ethics training

This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic.mit.edu/docs/gettingstarted/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

### Completion Report

<https://citiprogram.org/verify/?kde435228-50b9-4be9-b8df-b7afd56f2a99-46524805>

### Completion Certificate

<https://citiprogram.org/verify/?w0840cb69-a8c6-4caa-93d3-4975697f131d-46524805>

## Q3. Linux Shell Commands

1. The `/mnt/mimiciv/1.0` folder on teaching server contains data sets from MIMIC-IV. Refer to the documentation <https://mimic.mit.edu/docs/iv/> for details of data files.  
    ```{bash}
    ls -l /mnt/mimiciv/1.0
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/mnt/mimiciv/1.0` directly in following exercises. 

    Use Bash commands to answer following questions.

2. Display the contents in the folders `core`, `hosp`, `icu`. Why are these data files distributed as `.csv.gz` files instead of `.csv` (comma separated values) files? Read the page <https://mimic.mit.edu/docs/iv/> to understand what`s in each folder.

### Solution 
- Contents of the `core` folder are:

```{bash}
ls -l /mnt/mimiciv/1.0/core
```
- Contents of the `hosp` folder are:

```{bash}
ls -l /mnt/mimiciv/1.0/hosp
```
 - Contents of the `icu` folder are:
 
```{bash}
ls -l /mnt/mimiciv/1.0/icu
```
- The files are distributed as `.csv.gz` instead of `.csv` because the files are 
compressed by the gzip utility to reduce file size, allowing for faster file 
transfer.

3. Briefly describe what bash commands `zcat`, `zless`, `zmore`, and `zgrep` do.

### Solution 
<<<<<<< HEAD
Note that the main advantage of these commands is that they do not uncompress 
=======

- Note that the main advantage of these commands is that they do not uncompress 
<<<<<<< HEAD
>>>>>>> 755537b... Fixed answer formatting for hw1 Q3 (all parts) and fixed bash commands for Q3.5-3.6
the target files themselves.
=======
the target file themselves.
>>>>>>> a1f6bfd... Finalized hw2 by making minor edits to Q3-5

- `cat` prints the contents of a gzip-compressed file.

- `zless` is used to view the contents of a gzip-compressed file screen by 
screen. The spacebar can be used to move to the next page, and the mouse can 
also be used to scroll through the file. Press `q` to exit.

- `zmore` is similar to `zless` but is more limited in functionality. 
For example, mouse scrolling is disabled.

- `zgrep` prints all lines from a gzip-compressed file that match a specified 
expression.

4. What`s the output of following bash script?
    ```{bash, eval=T}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        ls -l $datafile
      done
    ```

### Solution

The output is a list of all gzip-compressed files from MIMIC-IV in the long
list format. 

Display the number of lines in each data file using a similar loop.

### Solution 
```{bash, eval=T}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        zcat  $datafile | wc -l
      done
```
5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `sort`, `uniq`, `wc`, and so on.)

### Solution

- The first few lines are:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | head 
```
- The number of rows in the data file is:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | wc -l 
```
- The number of unique patients in this data file is:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n+2 | awk -F, '{print $1}'\
| sort | uniq | wc -l
``` 

6. What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq -c`, `wc`, and so on.)

### Solution

- Admission type:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n+2 | awk -F, '{print $6}'\
| sort | uniq -c
```
- Admission location:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n+2 | awk -F, '{print $7}'\
| sort | uniq -c
```
- Insurance:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n+2 | awk -F, '{print $9}'\
| sort | uniq -c
```
- Ethnicity:
```{bash}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n+2 | awk -F, \
'{print $12}' | sort | uniq -c
```
## Q4. Who`s popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=T}
    wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    ```
Explain what `wget -nc` does. Do **not** put this text file `pg42671.txt` in Git. Complete the following loop to tabulate the number of times each of the four characters is mentioned using Linux commands.

### Solution 

`wget` is a command-line utility used to download files from the web. It is 
non-interactive, so it can work in the background. It supports HTTPS, HTTP, and
FTP, the most widely used Internet protocals. The `-nc` option prevents a file
from being downloaded again if it has already been downloaded.

```{bash, eval=T}
    #wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
    for char in Elizabeth Jane Lydia Darcy
    do
      echo $char:
       cat pg42671.txt | grep -oi $char | wc -l
    done
```

2. What`s the difference between the following two commands?
    ```{bash eval=FALSE}
    echo `hello, world` > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo `hello, world` >> test2.txt
    ```

### Solution 

Both the `>` and `>>` operators will create the specified file if it does not 
exist already in the current directory and add `hello, world` to it. 
However, if the files have already been created, the `>` operator will overwrite 
the existing file while `>>` will append to the existing file.

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=F}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=T}
    chmod 700 middle.sh
    ./middle.sh pg42671.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

### Solution

The output is lines 16-20 from the downloaded version of *Pride and Prejudice*.
`"$1"` is the first argument passed to the `middle.sh` script, `"$2"` is the
second argument passed to the script, and `"$3"` is the third argument. When 
executing the script above, `pg42671.txt` is the first argument, `20` is the 
second argument, and `5` is the third argument. Therefore,
the commands being executed above, according to the script is:
`head -n 20 pg42671.txt | tail -n 5`. The first line of the shell script 
specifies that the script should always be run with the Bourne Shell (sh).

### Q5. More fun with Linux

Try following commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.
<<<<<<< HEAD
=======
``` {bash}
   cal
   cal 2021
   cal 9 1752
   date
```   
```{bash}   
   hostname
   arch
   uname -a
   uptime
```

```{bash}
   who am i
   who
```
```{bash}
   w
```

```{bash}
   id
   last | head
```

```{bash}
   echo {con,pre}{sent,fer}{s,ed}
   time sleep 5
```

``` {bash}
history | tail
```

### Solution
- `cal` outputs a calendar of the current month with the current day 
highlighted.

- `cal 2021` outputs a calendar of the year 2021.

- `cal 9 1752` outputs a calendar of September 1752. September 3-13 are not 
shown on the calendar.

- `date` outputs the abbreviated day of the week, abbreviated month, day of 
the month, current time, the abbreviated time zone name, and the current year.

- `hostname` outputs the current hostname of the Linux system.

- `arch` outputs the current machine architecture. The teaching 
server runs a 64-bit implementation of the x86 architecture set, making the
architecture x86_64.

- `uname -a` outputs the operating system kernel name, the hostname, 
the kernel release, the last time the kernel was compiled in the format of the 
`date` command, the machine architecture, the processor architecture, 
the operating system architecture, and the current operating system.

- `uptime` outputs the current time in the form hr:min:sec, how long the 
system has been up in the format days, hr:min, the number of users currently 
logged on, and the load averages over the past 1, 5, and 15 minutes,  
which describe the average number of  processes that are currently being 
executed or waiting to be executed.

- `whoami` outputs the username, terminal, login time, and hostname of the 
current user.

- `who` outputs a list of all users currently logged on the system. The output 
includes: username, terminal, login time, and the hostname or IP address.

- `w` outputs the content from the `uptime` command as a head as well as a
list similar to the output of the `who` command but with more information
including idle time, JCPU time (time used by all processes on terminal), PCPU 
time (time used by current process), and command line of current process.

- `id` outputs the current user`s user id, primary group id (2x), supplementary
group ids (if they exist), and security context. 

- `last | head` displays information about the last ten login sessions. 
Information includes: username, terminal, IP address or hostname, time of login,
and duration of session (or `still logged in` if user hasn`t logged out).

- `echo {con,pre}{sent,fer}{s,ed}` prints all combinations of letters that start
with either `con` or `pre`, have either `ent` or `fer` in the middle, and end 
with `s` or `ed`. It just so happens that six out of the eight possible 
combinations are valid English words, with `confered` and `prefered` missing
the extra `r` to form the past tense.

- `time sleep 5` displays the real-world time elapsed, time 
spent in user mode, and time spent in kernel mode for a dummy job that is
delayed for 5 seconds.

- `history | tail` displays the last ten commands executed (does not have to be 
valid) and their corresponding line numbers.

- Note that `cal`, `date`, `uptime`, `who`, `w`, `last | head`, and 
`history |  tail` depend on time of execution, and `hostname`, `arch`, 
`uname -a`, `uptime`, `who`, `w`, `id`, `last | head`, and `history | tail` 
depends on Linux system, user, or both, so output presented in the submitted
html file will be very different from the output when this Rmd is re-knit.

- Both `who am i` and `history | tail` do not yield any output after knitting 
the Rmd file, so my answers were based on the output from my laptop's terminal.


>>>>>>> 87bd0d4... Finished preliminary draft of hw1
