---
title: "Biostat 203B Homework 2"
author: "Jonathan King"
subtitle: Due Feb 8 @ 11:59PM
output:
  html_document:
    toc: yes
    toc_depth: 4
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

```{r setup, message=F}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
# Set root directory to simplify file importing
knitr::opts_knit$set(root.dir = '/mnt/mimiciv/1.0')
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/mnt/mimiciv/1.0"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-1.0"
}
```

In this exercise, we use tidyverse (ggplot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2022winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

```{r}
# tree -s -L 2 /Users/huazhou/Documents/Box\ Sync/MIMIC/mimic-iv-1.0
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the tidyverse.

### Solution

```{r}
# Compare run times of different file reading functions
system.time(read_1 <- read.csv("./core/admissions.csv.gz"))
system.time(read_2 <- read_csv("./core/admissions.csv.gz"))
system.time(read_3 <- fread("./core/admissions.csv.gz"))
# Check for differences in parsed data types
str(read_1)
str(read_2)
str(read_3)
# Delete temporary reads
rm(read_1, read_2, read_3)
```

- `fread` is the fastest function overall, with it and 'read_csv'
both being much faster than `read.csv`. By default, `read.csv` parsed all character variables
and dates as factors, while both `read_csv` and `fread` correctly parsed character variables 
and parsed dates as POSIXct variables. Also, `read_csv` parsed id numbers as numeric, while the other
two file reading functions parsed the id numbers as integers.

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/icustays.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `icustays.csv.gz` as a tibble `icustays_tble`. 

### Solution

```{r}
icustays_tble <- read_csv("./icu/icustays.csv.gz") %>%
  print(width = Inf)
```

2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 

### Solution

```{r}
icustays_tble %>%
  group_by(subject_id) %>%
  summarise(number_of_stays = n_distinct(stay_id)) %>%
  arrange(desc(number_of_stays))

```

- There are 53,150 unique `subject_id`, and each one can have multiple ICU stays.

3. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`.

### Solution

```{r}
icustays_tble <- icustays_tble %>%
  group_by(subject_id) %>% 
  arrange(intime) %>% 
  distinct(subject_id, .keep_all = TRUE) %>%
  print(width = Inf)
```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/admissions.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.

### Solution

```{r}
admissions_tble <- read_csv("./core/admissions.csv.gz") %>%
  print(width = Inf)
```

2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadmi_id`.

### Solution

```{r}
admissions_tble <- admissions_tble %>%
  semi_join(icustays_tble, by = c("subject_id", "hadm_id")) %>%
  print(width = Inf)
```

3. Summarize the following variables by graphics. 

- admission year  
- admission month  
- admission month day  
- admission week day  
- admission hour (anything unusual?)  

### Solution

```{r}
# Summarize by admission year
admissions_tble %>% 
  mutate(year = year(admittime)) %>% 
  ggplot() +
  geom_histogram(mapping = aes(x = year), fill = "skyblue", binwidth = 15) +
  labs(title = " ICU Admissions by Year", x = "Year", y = "Admission Count") +
  theme(plot.title = element_text(hjust = 0.5))
# Summarize by admission month
admissions_tble %>% 
  mutate(month = month(admittime, label = TRUE)) %>% 
  ggplot() +
  geom_bar(mapping = aes(x = month, fill = month)) +
  scale_fill_brewer(palette = "Set3") + 
  theme(legend.position = "none") +
  labs(title = "ICU Admissions by Month", x = "Month", y = "Admission Count") +
  theme(plot.title = element_text(hjust = 0.5))
# Summarize by day of the month
admissions_tble %>% 
  mutate(mday = mday(admittime)) %>% 
  ggplot() +
  geom_histogram(
    mapping = aes(x = mday), 
    fill = "lightblue", 
    binwidth = 5,
    center = 2.5
  ) +
  labs(
    title = "ICU Admissions by Day of the Month", 
    x = "Day of the Month",
    y = "Admission Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
# Summarize by day of the week
admissions_tble %>% 
  mutate(wday = wday(admittime, label = TRUE)) %>% 
  ggplot() +
  geom_bar(mapping = aes(x = wday, fill = wday)) +
  theme(legend.position = "none") +
  labs(
    title = "ICU Admissions by Day of the Week", 
    x = "Day of the Week",
    y = "Admission Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
# Summarize by admission hour
admissions_tble %>% 
  mutate(hour = hour(admittime)) %>% 
  ggplot() +
  geom_bar(mapping = aes(x = hour)) +
  theme(legend.position = "none") +
  labs(title = "ICU Admissions by Hour", x = "Hour", y = "Admission Count") +
  theme(plot.title = element_text(hjust = 0.5))
```

- The 12 a.m. hour was when the most patients were admitted. Fewer patients
are admitted in the morning hours after, except for an unusual spike in the 7 a.m. hour.

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/patients.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

### Solution

```{r}
patients_tble <- read_csv("./core/patients.csv.gz") %>%
  semi_join(icustays_tble) %>%
  print(width = Inf)
```

2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.

### Solution

```{r}
# Make barplot of gender
ggplot(data = patients_tble) +
  geom_bar(aes(x = gender, fill = gender)) +
  theme(legend.position = "none") +
  labs(title = "Patient Gender", x = "Gender", y = "Patient Count") +
  theme(plot.title = element_text(hjust = 0.5))
# Create histogram of patient anchor age
ggplot(data = patients_tble) +
  geom_histogram(
    aes(x = anchor_age), 
    fill = "orangered2", 
    binwidth = 5,
    center = 17.5
  ) +
  labs(
    title = "Patient Anchor Age", 
    x = "Anchor Age (yrs.)", 
    y = "Patient Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
# Generate numeric summary of anchor_age by gender
patients_tble %>%
  group_by(gender) %>%
  summarise(
    n = n(), 
    mean_anchor_age = mean(anchor_age), 
    median_anchor_age = median(anchor_age), 
    sd_anchor_age = sd(anchor_age),
    min_anchor_age = min(anchor_age),
    max_anchor_age = max(anchor_age)
  ) %>%
  print(width = Inf)
```

- Most patients were male and had an anchor age above 60, with the highest number of patients having an anchor age between 65 and 70 compared to any other five-year age range.
Mean and median anchor age were lower for male patients compared to female patients. 
Standard deviation in anchor age was also lower for male patients.
Minimum patient anchor age was 18 years, and maximum patient anchor age was 91 years for both genders.

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/d_labitems.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.

### Solution

```{bash}
zcat /mnt/mimiciv/1.0/hosp/labevents.csv.gz | tail -n+2 | wc -l 
```

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), glucose (50931), magnesium (50960), and calcium (50893). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` and `lazy` options in the `read_csv` function to reduce the memory burden.

### Solution

```{r}
labevents_tble <- read_csv("./hosp/labevents_filtered_itemid.csv.gz") %>%
  semi_join(icustays_tble) %>%
  print()
# Read in dictionary of lab measurements
labevents_dict <- read_csv("./hosp/d_labitems.csv.gz")
```

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay.

### Solution

```{r} 
# Create table to determine first lab measurement
chartime <- select(icustays_tble, subject_id, intime)
# Find first measurement after check-in and restrict lab events
labevents_tble <- left_join(labevents_tble, chartime) %>% 
  arrange(subject_id) %>%
  filter(charttime >= intime) %>%
  distinct(subject_id, itemid, .keep_all = TRUE) %>%
# Create separate columns for each lab measurement and use measurement names
  inner_join(labevents_dict, by = "itemid") %>%
  select(subject_id, label, valuenum) %>%
  pivot_wider(names_from = label, values_from = valuenum) %>%
  print(width = Inf)
```

4. Summarize the lab measurements by appropriate numerics and graphics. 

### Solution

```{r}
# Calculate summary statistics for all lab measurements
labevents_tble %>%
  select(-subject_id) %>%
  summarize_all( 
    list(
      n_valid = ~ sum(!is.na(.)), 
      mean = mean,
      sd = sd,
      median = median,
      min = min,
      max = max
    ),
    na.rm = TRUE
  ) %>%
  print(width = Inf)
# Make a boxplot for each lab measurement unfiltered
for (i in 2:11) {
  # Temporary variables needed to make plots
  name <- colnames(labevents_tble[, i])
  col <- labevents_tble[, i]
  col_vec <- labevents_tble[[i]]
  iqr <- IQR(col_vec, na.rm = TRUE)
  q1 <- quantile(col_vec, 0.25, na.rm = TRUE)
  q3 <- quantile(col_vec, 0.75, na.rm = TRUE)
  sd <- sd(col_vec, na.rm = TRUE)
  # Constructing plots inside the loop
  print(
    labevents_tble %>%
      ggplot() +
      geom_boxplot(mapping = aes(y = .data[[name]])) +
      labs(title = paste("Patient", name, "Levels"), y = "Measurement Value") +
      theme(
        plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
      )
  )
}
# Make a boxplot for each lab measurement filtered using the 1.5 IQR rule
for (i in 2:11) {
  # Temporary variables needed to make plots
  name <- colnames(labevents_tble[, i])
  col <- labevents_tble[, i]
  col_vec <- labevents_tble[[i]]
  iqr <- IQR(col_vec, na.rm = TRUE)
  q1 <- quantile(col_vec, 0.25, na.rm = TRUE)
  q3 <- quantile(col_vec, 0.75, na.rm = TRUE)
  sd <- sd(col_vec, na.rm = TRUE)
  # Constructing plots inside the loop
  print(
    labevents_tble %>%
      filter(col < (q3 + 1.5 * iqr) & col > (q1 - 1.5 * iqr)) %>%
      ggplot() +
      geom_boxplot(mapping = aes(y = .data[[name]])) +
      labs(title = paste("Patient", name, "Levels"), y = "Measurement Value") +
      theme(
        plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
      )
  )
}
# Remove temporary variables
rm(name, col, col_vec, iqr, sd, q1, q3)
```

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/chartevents.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/d_items.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

### Solution

```{r}
chartevents_tble <- read_csv("./icu/chartevents_filtered_itemid.csv.gz") %>% 
  semi_join(icustays_tble, by = "subject_id") %>%
  print()
# Read in dictionary of chart measurements
 chartevents_dict <- read_csv("./icu/d_items.csv.gz")
 

```
2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 

### Solution

```{r}
# Find first measurement after check-in and restrict chart events
chartevents_tble <- left_join(chartevents_tble, chartime) %>%
  filter(charttime >= intime) %>%
  distinct(subject_id, itemid, .keep_all = TRUE) %>%
  arrange(subject_id) %>%
  # Create separate columns for each vital measurement and use measurement names
  inner_join(chartevents_dict, by = "itemid") %>%
  select(subject_id, label, valuenum) %>%
  pivot_wider(names_from = label, values_from = valuenum) %>%
  print(width = Inf)
```
3. Summarize these vital measurements by appropriate numerics and graphics. 

### Solution

```{r}
# Summarize all vital measurements
chartevents_tble %>% 
  select(-subject_id) %>%
  summarize_all( 
    list(
      n_valid = ~ sum(!is.na(.)), 
      mean = mean,
      sd = sd,
      median = median,
      min = min,
      max = max
    ),
    na.rm = TRUE
  ) %>%
  print(width = Inf)
# Make boxplots for vital measurements unfiltered
for (i in 2:6)
{
  # Temporary variables needed to make plots
  name <- colnames(chartevents_tble[, i])
  col <- chartevents_tble[, i]
  col_vec <- chartevents_tble[[i]]
  iqr <- IQR(col_vec, na.rm = TRUE)
  q1 <- quantile(col_vec, 0.25, na.rm = TRUE)
  q3 <- quantile(col_vec, 0.75, na.rm = TRUE)
  sd <- sd(col_vec, na.rm = TRUE)
  # Constructing plots inside the loop
  print(
    chartevents_tble %>%
      ggplot() +
      geom_boxplot(mapping = aes(y = .data[[name]])) +
      labs(title = paste("Patient", name), y = "Measurement Value") +
      theme(
        plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
      )
  )
}
# Make a boxplot for each vital measurement filtered using the 1.5 IQR rule
for (i in 2:6)
{
  # Temporary variables needed to make plots
  name <- colnames(chartevents_tble[, i])
  col <- chartevents_tble[, i]
  col_vec <- chartevents_tble[[i]]
  iqr <- IQR(col_vec, na.rm = TRUE)
  q1 <- quantile(col_vec, 0.25, na.rm = TRUE)
  q3 <- quantile(col_vec, 0.75, na.rm = TRUE)
  sd <- sd(col_vec, na.rm = TRUE)
  # Constructing plots inside the loop
  print(
    chartevents_tble %>%
      filter(col < ( q3 + 1.5 * iqr) & col > (q1 - 1.5 * iqr)) %>%
      ggplot() +
      geom_boxplot(mapping = aes(y = .data[[name]])) +
      labs(title = paste("Patient", name), y = "Measurement Value") +
      theme(
        plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
      )
  )
}
# Remove temporary variables
rm(name, col, col_vec, iqr, sd, q1, q3)
# Remove temporary table need to determine first measurements
rm(chartime)
```

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are  

- first ICU stay of each unique adult (age at admission > 18)

and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

### Solution

```{r}
# Create 30 day mortality variable
admissions_tble <- admissions_tble %>%
  mutate(thirty_day_mort = ifelse(deathtime - admittime < 43200, "Yes", "No"))
# Assign 30 day mortality as "No" for individuals who did not pass away
admissions_tble$thirty_day_mort[is.na(admissions_tble$thirty_day_mort)] <- "No"
mimic_icu_cohort <- admissions_tble %>%
  inner_join(icustays_tble, by = "subject_id", "hadm_id") %>%
  inner_join(patients_tble) %>%
  left_join(labevents_tble) %>%
  left_join(chartevents_tble) %>%
# Restrict age at admission > 18
  filter(year(admittime) - (anchor_year - anchor_age) > 18) %>%
  print(width = Inf)
  
```
## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

### Solution

```{r}
# Make ethnicity chart
mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = thirty_day_mort)) +
  theme(legend.position = "none") +
  labs(title = "30 Day Mortality by Ethnicity", x = "30 Day Mortality") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(
    ~ ethnicity, 
    scales = "free", 
    nrow = 3, 
    labeller = labeller(ethnicity = label_wrap_gen(width = 20))
  )
# Make language chart
mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = thirty_day_mort)) +
  theme(legend.position = "none") +
  labs(title = "30 Day Mortality by Language", x = "30 Day Mortality") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ language, scales = "free", nrow = 2)
# Make insurance chart
mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = thirty_day_mort)) +
  theme(legend.position = "none") +
  labs(title = "30 Day Mortality by Insurance", x = "30 Day Mortality") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ insurance, scales = "free", nrow = 3)
# Make maritial status chart
mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = thirty_day_mort)) +
  theme(legend.position = "none") +
  labs(title = "30 Day Mortality by Marital Status", x = "30 Day Mortality") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ marital_status, scales = "free", nrow = 3)
# Make gender chart
mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = thirty_day_mort)) +
  theme(legend.position = "none") +
  labs(title = "30 Day Mortality by Gender", x = "30 Day Mortality") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ gender, scales = "free", nrow = 2)

# Make age at admission plot
mimic_icu_cohort %>%
  mutate(age = (year(admittime) - (anchor_year - anchor_age))) %>%
  ggplot() +
  geom_boxplot(
    mapping = aes(
    x = thirty_day_mort, 
    y = age, 
    fill = thirty_day_mort)
  ) +
  theme(legend.position = "none") +
  labs(
    title = "Age at Admission by 30 Day Mortality", 
    y = "Age (yrs.)",
    x = "30 Day Mortality"
  ) +
  theme(plot.title = element_text(hjust = 0.5)) 
  
  
```

- `thirty_day_mort` vs first lab measurements

### Solution

```{r}
mimic_icu_cohort %>%
  group_by(thirty_day_mort) %>%
  select(16, 29:38) %>%
  summarize_all( 
    list(
      n_valid = ~ sum(!is.na(.)), 
      mean = mean,
      sd = sd,
      median = median,
      min = min,
      max = max
    ),
    na.rm = TRUE
  ) %>%
  print(width = Inf)
```

### Solution

- `thirty_day_mort` vs first vital measurements

```{r}
mimic_icu_cohort %>%
  group_by(thirty_day_mort) %>%
  select(16, 39:43) %>%
  summarize_all( 
    list(
      n = ~ sum(!is.na(.)), 
      mean = mean,
      sd = sd,
      median = median,
      min = min,
      max = max
    ),
    na.rm = TRUE
  ) %>%
  print(width = Inf)
```

- `thirty_day_mort` vs first ICU unit

### Solution

```{r}
mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = thirty_day_mort)) +
  theme(legend.position = "none") +
  labs(title = "30 Day Mortality by First ICU Unit ", x = "30 Day Mortality") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(
    ~ first_careunit, 
    scales = "free", 
    nrow = 3, 
    labeller = labeller(first_careunit = label_wrap_gen(width = 25))
  )
```

